Semi-Parametric Manifold Clustering

### 2021-12-08 curve-fitting-methods/curve-fitting-scratch.Rmd
updated notes

where $T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}$ and each $t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r$, and $X \in \mathbb{R}^{n \times d}$. 
The same procedure for estimating $t_1, ..., t_n$ can be applied here.

@@

>>>>

where $T = \begin{bmatrix} t_{\cdot, 1} & \cdots & t_{\cdot, R} \end{bmatrix}$ and each $t_{ir} = \binom{R}{r} (1 - t_i)^{R - r} t_i^r$, and $X \in \mathbb{R}^{n \times d}$. 
The same procedure for estimating $t_1, ..., t_n$ can be applied here.

@@

## Theoretical Framework

If we assume that the points are distributed as $X_i \ind \mathcal{N}(f(t_i | p), \Sigma(t_i | \phi))$, then we can write the (incomplete) log likelihood as:

$$\ell(p, \phi) = -\frac{1}{2} \sum_i \log |\Sigma(t_i | \phi)| - \frac{1}{2} \sum_i (x_i - f(t_i | p))^\top (\Sigma(t_i | \phi))^{-1} (x_i - f(t_i | p))$$

In the case $\Sigma(t_i | \phi) = \phi I$ and $f(t_i | p) = p^\top \tilde{t}_i$ where $\tilde{t}_i \in \mathbb{R}^{R+1}$ is the Bezier polynomial expansion of order $R$, then this becomes

$$\ell(p, \phi) = -\frac{n d}{2} \log \phi - \frac{1}{2 \phi} \|X - T p\|^2_F$$

Then given $t_1, ..., t_n$, the MLE of $p$ is again $(T^\top T)^{-1} T^\top X$, and the MLE of $\phi$ is just $\frac{1}{n d} \|X - T \hat{p}\|^2_F$.
And given $p$, we can estimate each $t_i$ in the same way as before to maximize $\ell$ (the variance is unnecessary here).

As noted before, the parameterization given by $T$ and $p$ is not unique. 
We can arbitrarily scale the $t_i$'s and $p$ to obtain the same value of $\ell$. 
Naively performing the proposed algorithm sometimes results in estimates for $p$ that diverge and estimates for $t_i$'s that converge to a single value. 
To remedy this, we can either scale the $t_i$'s as before, 
or we can scale $p$ by forcing $f$ to be an arclength parameterization.

\begin{theorem}
Let $x_1, ..., x_n \ind \mathcal{N}(p^\top \tilde{t}_i, \phi I_d)$,
where each $x_i \in \mathbb{R}^d$, $\tilde{t}_i \in \mathbb{R}^{R+1}$ such that $\tilde{t}_{i,r} = \binom{R}{r} (1 - t_i)^{R-r} t_i^r$ for $i = 1, ..., n$ and $r = 0, ..., R$, 
and $p \in \mathbb{R}^{(R+1) \times d}$.

Then each iteration of the following decreases the negative log likelihood:

\begin{enumerate}
  \item $p^{(s)} \leftarrow (T^\top T)^{-1} T^\top X$, 
where $T = \begin{bmatrix} \tilde{t}_1 & \cdots & \tilde{t}_n \end{bmatrix}^\top$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
  \item $t_i^{(s)} \leftarrow \arg\min_t \|x_i - p^\top \tilde{t}_i\|^2$
\end{enumerate}
\end{theorem}

\begin{proof}[Proof (sketch)]

For $\Sigma(t_i | \phi) = \phi I$, the negative log likelihood is, 
up to some additive and multiplicative constants:

$$l(p, \{t_i\}) = \|X - T p\|^2_F = X^\top X - 2 X^\top T p + p^\top T^\top T p$$

To find the minimizer $\hat{p}$ given $t_1, ..., t_n$:

$\nabla_p l = -2 T^\top X + 2 T^\top T p = 0$  
$\implies \hat{p} = (T^\top T)^{-1} T^\top X$

To find the minimizer $\hat{t}_i$ given $p$:

We can rewrite the negative log likelihood as 
$l(p, \{t_i\}) = \sum_i \|x_i - p^\top \tilde{t}_i\|^2$.

Then each entry of $\nabla_{t} l$ depends only on $t_i$, so each $t_i$ can be optimized independently. 
Furthermore, each $\frac{\partial}{\partial t_i} \|x_i - p^\top \tilde{t}_i\|^2$ is a polynomial in $t_i$, so it can be minimized by simply finding the roots of the polynomial.

Thus this algorithm is a coordinate descent algorithm, which reduces the objective function with each iteration. 
\end{proof}

\begin{theorem}

Let $A^{(n)} \sim \RDPG(F(p, \theta), n)$, and let $X^{(n)}$ be the ASE of $A^{(n)}$. 
Let $\hat{p}, \{\hat{t}_i\}$ be the global minimizer of $l$ given $X^{(n)}$. 
Then $l(\hat{p}, \{\hat{t}_i\}) \stackrel{a.s.}{\to} 0$.

\end{theorem}

### Covariance Structures

While the CLT property of the ASE allows us to approximate $x_i \sim \mathcal{N} (W f(t_i | p), W \Sigma(t_i | \phi) W^\top)$, it is unclear what $\Sigma(t_i | \phi)$ looks like, and we cannot in general say $\Sigma(t_i | \phi) = \phi I_d$.

### 2021-12-08 curve-fitting-methods/curve-fitting-scratch.tex
updated notes

where \(T = \begin{bmatrix} t^{(1)} & \cdots & t^{(R)} \end{bmatrix}\)
and each \(t^{(r)}_i = \binom{R}{r} (1 - t_i)^{R - r} t_i^r\), and
\(X \in \mathbb{R}^{n \times d}\). The same procedure for estimating
\(t_1, ..., t_n\) can be applied here.

@@

>>>>

where
\(T = \begin{bmatrix} t_{\cdot, 1} & \cdots & t_{\cdot, R} \end{bmatrix}\)
and each \(t_{ir} = \binom{R}{r} (1 - t_i)^{R - r} t_i^r\), and
\(X \in \mathbb{R}^{n \times d}\). The same procedure for estimating
\(t_1, ..., t_n\) can be applied here.

@@

\hypertarget{theoretical-framework}{%
\subsection{Theoretical Framework}\label{theoretical-framework}}

If we assume that the points are distributed as
\(X_i \stackrel{\mathrm{ind}}{\sim}\mathcal{N}(f(t_i | p), \Sigma(t_i | \phi))\),
then we can write the (incomplete) log likelihood as:

\[\ell(p, \phi) = -\frac{1}{2} \sum_i \log |\Sigma(t_i | \phi)| - \frac{1}{2} \sum_i (x_i - f(t_i | p))^\top (\Sigma(t_i | \phi))^{-1} (x_i - f(t_i | p))\]

In the case \(\Sigma(t_i | \phi) = \phi I\) and
\(f(t_i | p) = p^\top \tilde{t}_i\) where
\(\tilde{t}_i \in \mathbb{R}^{R+1}\) is the Bezier polynomial expansion
of order \(R\), then this becomes

\[\ell(p, \phi) = -\frac{n d}{2} \log \phi - \frac{1}{2 \phi} \|X - T p\|^2_F\]

Then given \(t_1, ..., t_n\), the MLE of \(p\) is again
\((T^\top T)^{-1} T^\top X\), and the MLE of \(\phi\) is just
\(\frac{1}{n d} \|X - T \hat{p}\|^2_F\). And given \(p\), we can
estimate each \(t_i\) in the same way as before to maximize \(\ell\)
(the variance is unnecessary here).

As noted before, the parameterization given by \(T\) and \(p\) is not
unique. We can arbitrarily scale the \(t_i\)'s and \(p\) to obtain the
same value of \(\ell\). Naively performing the proposed algorithm
sometimes results in estimates for \(p\) that diverge and estimates for
\(t_i\)'s that converge to a single value. To remedy this, we can either
scale the \(t_i\)'s as before, or we can scale \(p\) by forcing \(f\) to
be an arclength parameterization.

\begin{theorem}
Let $x_1, ..., x_n \stackrel{\mathrm{ind}}{\sim}\mathcal{N}(p^\top \tilde{t}_i, \phi I_d)$,
where each $x_i \in \mathbb{R}^d$, $\tilde{t}_i \in \mathbb{R}^{R+1}$ such that $\tilde{t}_{i,r} = \binom{R}{r} (1 - t_i)^{R-r} t_i^r$ for $i = 1, ..., n$ and $r = 0, ..., R$, 
and $p \in \mathbb{R}^{(R+1) \times d}$.

Then each iteration of the following decreases the negative log likelihood:

\begin{enumerate}
  \item $p^{(s)} \leftarrow (T^\top T)^{-1} T^\top X$, 
where $T = \begin{bmatrix} \tilde{t}_1 & \cdots & \tilde{t}_n \end{bmatrix}^\top$ and 
$X = \begin{bmatrix} x_1 & \cdots & x_n \end{bmatrix}^\top$.
  \item $t_i^{(s)} \leftarrow \arg\min_t \|x_i - p^\top \tilde{t}_i\|^2$
\end{enumerate}
\end{theorem}

\begin{proof}[Proof (sketch)]

For $\Sigma(t_i | \phi) = \phi I$, the negative log likelihood is, 
up to some additive and multiplicative constants:

$$l(p, \{t_i\}) = \|X - T p\|^2_F = X^\top X - 2 X^\top T p + p^\top T^\top T p$$

To find the minimizer $\hat{p}$ given $t_1, ..., t_n$:

$\nabla_p l = -2 T^\top X + 2 T^\top T p = 0$  
$\implies \hat{p} = (T^\top T)^{-1} T^\top X$

To find the minimizer $\hat{t}_i$ given $p$:

We can rewrite the negative log likelihood as 
$l(p, \{t_i\}) = \sum_i \|x_i - p^\top \tilde{t}_i\|^2$.

Then each entry of $\nabla_{t} l$ depends only on $t_i$, so each $t_i$ can be optimized independently. 
Furthermore, each $\frac{\partial}{\partial t_i} \|x_i - p^\top \tilde{t}_i\|^2$ is a polynomial in $t_i$, so it can be minimized by simply finding the roots of the polynomial.

Thus this algorithm is a coordinate descent algorithm, which reduces the objective function with each iteration. 
\end{proof}

\begin{theorem}

Let $A^{(n)} \sim \mathrm{RDPG}(F(p, \theta), n)$, and let $X^{(n)}$ be the ASE of $A^{(n)}$. 
Let $\hat{p}, \{\hat{t}_i\}$ be the global minimizer of $l$ given $X^{(n)}$. 
Then $l(\hat{p}, \{\hat{t}_i\}) \stackrel{a.s.}{\to} 0$.

\end{theorem}

\hypertarget{covariance-structures}{%
\subsubsection{Covariance Structures}\label{covariance-structures}}

While the CLT property of the ASE allows us to approximate
\(x_i \sim \mathcal{N} (W f(t_i | p), W \Sigma(t_i | \phi) W^\top)\), it
is unclear what \(\Sigma(t_i | \phi)\) looks like, and we cannot in
general say \(\Sigma(t_i | \phi) = \phi I_d\).
